---
title: "HW2"
author: "Jackson Dial"
date: '2022-09-27'
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

*In this homework, the objectives are to*

1. Work with dates in R and exploratory data visualization

2. Data Preprocessing and Transformation

Assignments will only be accepted in electronic format in RMarkdown (.rmd) files and knitted .html files. Please make sure to **print your knitted .html file into a pdf** before you submit it to the Gradescope, and you may only submit your .rmd file to Sakai. Your code should be adequately commented to clearly explain the steps you used to produce the analyses. RMarkdown homework files should be uploaded to Sakai with the naming convention date_lastname_firstname_HW[X].Rmd. For example, my second homework assignment would be named 20220922_Dunn_Jessilyn_HW2.Rmd. **It is important to note that 5 points will be deducted for every assignment that is named improperly.** Please add your answer to each question directly after the question prompt in the homework .Rmd file template provided below.

---

```{r message=FALSE}
library(tidyverse)
library(ggplot2)
library(lubridate)
library(patchwork)
library(gridExtra)
library(psych)
library(corrplot)
```

---

## Working with Dates and EDA (25 points)

1. From the stroke dataset in the file "healthcare-dataset-stroke-data.csv", create a new dataframe named "stroke_df" that only contains information on the subjects' age, gender, hypertension, heart_disease, work type, BMI, smoking status, average glucose level (avg_glucose_level) as well as whether or not the patient had a stroke. 

```{r}
dat1 <- read_csv("Data/healthcare-dataset-stroke-data.csv")
stroke_df <- dat1 %>% 
  select(age, gender, hypertension, heart_disease, work_type, bmi, smoking_status, avg_glucose_level, stroke)
```

2. Cast all columns into the most sensible variable type, e.g., "age" should be an integer, "gender" should be a factor, "stroke" should be a logical, etc. 

```{r}
#There are some values for age that are decimal values, but the as.integer() function
#will truncate them, not round them because it is not true that someone is 15 if he/she
#is 14.9
stroke_df$age <- as.integer(stroke_df$age)
stroke_df$gender <- as.factor(stroke_df$gender)
stroke_df$hypertension <- as.factor(stroke_df$hypertension)
stroke_df$heart_disease <- as.factor(stroke_df$heart_disease)
stroke_df$work_type <- as.factor(stroke_df$work_type)
stroke_df$bmi <- as.numeric(stroke_df$bmi)
stroke_df$smoking_status <- as.factor(stroke_df$smoking_status)
stroke_df$avg_glucose_level <- as.numeric(stroke_df$avg_glucose_level)
stroke_df$stroke <- as.factor(stroke_df$stroke)
```


3. Using ggplot and the grid drawing method of your choice, for each stroke status, draw a density histogram with the density curve overlaid on the same plot, and wrap the plots in a 1-by-2 grid. Do this for **avg_glucose_level** and **bmi**. It is recommended to use the library "patchwork" or “gridExtra” for making the 2-plot grid. Label your axes appropriately, clearly show the legends, and include a plot title for full credit. You should also choose a suitable histogram bin width for each variable and explain why you made that choice. 

Before plotting, use the following code to delete the missing terms in the column "BMI".

```{r}
stroke_df <- na.omit(stroke_df)

ggplot(stroke_df, aes(x = avg_glucose_level))+
  geom_histogram(aes(y = ..density..))+
  geom_density(col = "springgreen3", size = 1.3)+
  theme(panel.grid.minor = element_blank())+
  facet_grid(~stroke)+
  labs(x = "Average Glucose Level",
       y = "Density",
       title = "Distribution of Glucose Across Stroke Levels",
       subtitle = "0 = 'No Stroke', 1 = 'Stroke'")
#The default bin width of 30 was deemed satisfactory for the glucose plot
#because the density curve follows the height of the histogram rather well
#overall.

ggplot(stroke_df, aes(x = bmi))+
  geom_histogram(aes(y = ..density..), bins = 40)+
  geom_density(col = "dodgerblue", size = .9)+
  theme(panel.grid.minor = element_blank())+
  facet_grid(~stroke)+
  labs(x = "BMI",
       y = "Density",
       title = "Distribution of BMI Across Stroke Levels",
       subtitle = "0 = 'No Stroke', 1 = 'Stroke'")
#I chose a to change the number of bins to 40 to better fit the
#density distribution curve
  
```


4. What are your observations? E.g. Does any feature visually show a strong difference in means or in shape of the data between the stroke groups? For each feature, which stroke group seems to have a higher average value? Clearly explain your answer to receive full credit. 

The distribution of average glucose level across stroke groups does appear to differ. Though both plots show a bit of a bi-modal distribution, the stroke = 1 group shows a more distinct bi-modal distribution. The stroke = 1 group appears to have a larger average value.

The distribution of BMI across stroke levels is rather similar, though there is slightly more positive kurtosis in the stroke = 1 group. The stroke = 1 group appears to have a larger average value.

---

## Exploratory Data Analysis and Data Preprocessing (58 points)

5. Use ggplot and facet_wrap() to plot a bar plot for the following categorical variables in this dataset:

work_type
smoking_status

```{r}
ggplot(stroke_df, aes(y = work_type))+
  geom_bar()+
  facet_wrap(~smoking_status, nrow = 2)+
  theme(panel.grid.minor = element_blank())+
  labs(x = "Count",
       y = "Type of Work",
       title = "Count of Individuals Across Smoking and Work Statuses")

# theme(axis.text.x = element_text(angle = 90))+

```


+ When using facet_wrap(), you are encouraged to make 2 plots in one row. 
+ make sure to rotate the axis labels so that they are readable.


6. Using ggplot, draw density function curves of *age* by stroke status on the same plot. What do you observe? Which stroke group has a higher mean age?

+ remember to make a title, clearly label the axes, and create a legend.

```{r}
ggplot(stroke_df, aes(x = age, fill = stroke))+
  geom_density(alpha = .3)+
  theme(panel.grid.minor = element_blank())+
  labs(x = "Age",
       y = "Density",
       title = "Distribution of Age Across Stroke",
       subtitle = "0 = 'No Stroke', 1 = 'Stroke'")
```



7. Answer the following questions:

+ What is the average age of the subjects in this dataset who have had a stroke?
+ What is the average age of the subjects in this dataset who have not had a stroke?

+ What is the average BMI (body mass index) of subjects in this dataset who had heart_disease?
+ What is the average glucose level of subjects in this dataset who had heart_disease?

+ Among the  population who never smoked, what is the occurrence of stroke?
+ Among the population who did smoke, what is the occurrence of stroke?


```{r}
library(sjstats)
library(pander)

stroke_df %>%
  filter(stroke == 1) %>% 
  summarise(avg_age = mean(age)) %>% 
  pander()

stroke_df %>%
  filter(stroke == 0) %>% 
  summarise(avg_age = mean(age)) %>% 
  pander()

stroke_df %>%
  filter(heart_disease == 1) %>% 
  summarise(avg_bmi = mean(bmi)) %>% 
  pander()

stroke_df %>%
  filter(heart_disease == 0) %>% 
  summarise(avg_bmi = mean(bmi)) %>% 
  pander()

stroke_df %>% 
  filter(smoking_status == "never smoked") %>% 
  prop(stroke == 1)
  
stroke_df %>% 
  filter(smoking_status != "never smoked") %>% 
  prop(stroke == 1)

```


8. Now, using a similar 1 by 2 grid, overlay two separate density curves for the two groups based on their heart disease status, for each variable *BMI* and *average glucose level*.  For these figures, plot just the density curve (without the density histogram bins). Add a vertical line for each of the two density curves on each plot at the mean value for that feature for each group (grouped by heart disease status). For example, in the first entry of the grid, which is row 1 and column 1 of the 1-by-2 grid, there should be two density plots drawn in two colors based on whether subjects represented on the curve had heart disease or not.

```{r}
library(patchwork)

p1 <- ggplot(stroke_df, aes(x = bmi, fill = heart_disease))+
  geom_density(alpha = .3)+
  theme(panel.grid.minor = element_blank())+
  labs(x = "BMI",
       y = "Density",
       title = "Distribution of BMI Across Heart Disease")+
  theme(legend.position = "bottom",
        plot.title = element_text(size = 10))
p2 <- ggplot(stroke_df, aes(x = avg_glucose_level, fill = heart_disease))+
  geom_density(alpha = .3)+
  theme(panel.grid.minor = element_blank())+
  labs(x = "Average Glucose Level",
       y = "Density",
       title = "Distribution of Mean Glucose Across Heart Disease")+
  theme(legend.position = "bottom",
        plot.title = element_text(size = 10))

p1 + p2

# library(gridExtra)
# grid.arrange(p1,p2, nrow = 1, name = "test")
```


9. What are your observations from the previous plots? E.g. Do any of the features visually show a strong difference in means or in shape of the data distribution between the people who had heart disease and those who didn't? For each feature, which heart disease diagnosis group seems to have the higher average value? Clearly explain your answer to receive full credit.



The approximate averages of BMI appear to be similar across heart disease groups, but the averages of glucose appear to be different across the heart disease groups.

The average BMI appears to be higher in the heart disease = 1 group, because the peak of the blue curve is at a greater value than the red curve. 

The average glucose level appears to be higher in the heart disease = 1 group, because the peak on the right is much larger in the blue group than the red group.


10. Plot the histogram and density plots of the variable "avg_glucose_level", and answer the following questions:

```{r}
ggplot(stroke_df, aes(avg_glucose_level))+
  geom_histogram(aes(y = ..density..), bins = 40)+
  geom_density(col = "forestgreen", size = 2)+
  labs(x = "Average Glucose Level",
       y = "Density",
       title = "Distribution of Average Glucose Level")+
  theme(panel.grid.minor = element_blank())
```


a. Is the data skewed?
__Yes, the data is skewed.__
b. Does it have positive skewness or negative skewness? 
__The data has positive skewness.__
c. Compute the skewness using the definition from the lecture.
```{r}
n <- nrow(stroke_df)
x_bar = mean(stroke_df$avg_glucose_level)
gluc <- as.vector(stroke_df$avg_glucose_level)
skew <- ((1 / n) * sum((gluc - x_bar)^3)  ) / ((1 / (n - 1)) * sum((gluc - x_bar)^2) )^ (3/2)
skew
```

d. According to the criterion introduced in the lecture, is the dataset moderately skewed or highly skewed?


11. Correlation plots are a way to visualize multivariate relationships (between numerical variables). Using the corrplot package, make a correlation plot of the variables age, BMI and avg_glucose_level. Clearly label your axis and include a legend to receive full credit. 
(you should also add a title of the overall plot and coefficient values in each cell).



12. Calculate the z-scores of the *avg_glucose_level* and determine if there are any outliers using the definition of z-score > 3. Plot a histogram of this variable (i.e., plot histogram of avg_glucose_level). Remove the outliers (as defined by z-score>3) and plot the histogram again. What difference do you notice between the two histograms?


13. Perform winsorization on the "BMI" variable. This is a technique that was not covered in lecture, but is another type of transformation that limits the impact of outliers in your analyses. Typically, you can decide upon a threshold (often, a specific range of percentiles) and replace all of the data points outside of the threshold with the closest value from within the threshold. An example from Wikipedia may be a helpful demonstration: “a 90% winsorization would see all data below the 5th percentile set to the 5th percentile, and data above the 95th percentile set to the 95th percentile.” Conduct a 90% winsorization on the "BMI" variable and plot a histogram of the winsorized data. What is your observation of the data distribution before and after winsorization? 


